Wind Veer and Speed in Turbulent Ekman Flow Part I: Scaling analysis and universal profile model

### 2024-11-13 write1/manuscript.tex
minor edits based on suggestion from Hauk; added cover letter

The Coriolis force bends the path of motion on a rotating sphere and establishes geostrophic equilibrium when in balance with a pressure gradient force. Wind veer away from the wind direction in geostrophic equilibrium is
(i) due to direct frictional effects in the very vicinity of the surface and
(ii) due to turbulence which exerts indirect frictional effects; these effects
cause a slow-down of the mean wind reducing the Coriolis force thus turning the wind
in favor of the pressure gradient force. 
%%
Not only does the veering set the frame of reference for surface layer theory, it also has
effects at small and large scales from large-scale dispersion via plume spreading to
cyclone spin-down \citep{svensson:BM2009}
and on the capabilities of data assimilation and accuracy of surface flux estimates \citep{brown:QJR2005}. 
%
From a large-scale perspective, the veering of wind across the planetary boundary layer determines
the amount of cross-isobaric mass-flux, commonly referred to as 'Ekman pumping' \citep{ekman:AMA1905},
and it is thus a key factor in the life-cycle of large-scale synoptic systems.
%
% %% This shear is probably not related to Ekman-turning 
% On the meso-scale, shear in the surface layer a crucial factor for the initiation and sustained strong convective events (REFERENCE?).
% 
Within the atmospheric boundary layer (ABL), directional shear of the wind in
the upper part of the surface layer may cause a systematic yaw for tall wind power generation devices
where blades reach into the Ekman layer, i.e. that part of the boundary layer where the wind starts to turn;
an exact estimate of such effects is critical in the site assessments for wind farms
\citep{calaf:PF2010, mirocha:WES2018}. 
%
\par
%
In the planetary boundary layer, wind veer is characterized by the surface veering angle $\alpha$
defined as the angle between the negative surface shear stress $\tau_\SFC$ and the geostrophic wind.
%
Surface veering $\alpha$ and geostrophic drag $Z\equiv u_\star/G$, where the friction velocity $u_\star\equiv\sqrt{|\tau_\SFC|/\rho}$,
uniquely determine the surface drag $\tau_\SFC$ in a turbulent Ekman flow. 
%
In any quantitative description of the surface layer, the friction velocity $u_\star$ is the dynamic scale
and $\alpha$ defines the horizontal alignment of the frame of reference, i.e. the rotation of the surface 
friction with respect to the outer wind direction. 
%
Knowledge about $u_\star$ and $\alpha$ is thus a prerequisite for any quantitative theory of the surface layer, 
and \citet{rossby:PIP1935} constrained the two parameters based on integral relations in the ABL.
%
Asymptotic similarity theory was later used by \cite{tennekes:JAS1973, blackadar:JAS1968},
and--based on his seminal direct numerical simulations (DNS) of Ekman flow--,~\cite{spalart:JFM1989} suggested
a modification to take into account effects of low to intermediate Reynolds numbers.
Later on, constants were re-evaluated with a focus on the ABL based on
observations \citep{hogstrom:BM1988,hogstrom:BM1996}
and numerical modelling \citep{spalart:PF2008, spalart:PF2009, ansorge:BM2014, ansorge:BM2019}.
%
\par
%
Attempts were also undertaken to obtain profiles of the wind speed: 
One approach is to match the inner and outer layer at a reference height as usggested by
\cite{etling:2002, emeis:2018}  (Sec. 21.10; Eq. 21.48); they choose the Prandtl-layer height $z_\text{Prandtl}$ to match
the wind speed profiles, which, however, requires external prescription of $\alpha(z_\text{prandtl})$, the veering at that height. 
A one-dimensional profile with constant veering is given by \citet[Sec. 3; Eq. 3.1-3.19]{emeis:m2007}. 
%%%

@@

We therefore resort to the physical manifestation of the eddy diffuysivity in an Ekman layer, and use the

@@

This can, in particular for low or extremely high $\RE$ cause discontinuity in the derivatives of the velocity profiles around the transition from the inner to the outer layer. 

@@

height $z_{blend}^-=0.28-2.25\sqrt{1./Re_\tau}$ and a transition thickness 

@@

\[
  \omega(z^-) = \frac12 \left[\text{erf}\left(z_\text{trans}\log\frac{z^-}{z_{\text{blend}}^{-}}\right) +1\right].
\]

@@

%
\begin{figure}
  \phantom{AAA}\textbf{(a)\hspace{0.3\textwidth}(b)\hspace{0.3\textwidth}(c)}\\
  \includegraphics[width=\textwidth]{../plot/ekman_profiles.pdf}
  \caption{ \textbf{(a)} Velocity deficit, \textbf{(b)} velocity profile in shear-aligned hodographs and \textbf{(c)} hodograph in geostrophy-aligned coordinates.
   In panels \textbf{(a)} and \textbf{(b)}, blue lines correspond to the streamwise component and red to the spanwise. }
  \label{fig:summary} 
\end{figure} 
The resulting velocity profiles across the entire boundary layer are shown in inner and outer scaling as well as in ho\-do\-graph- view in Fig.~\ref{fig:summary}. The shear-aligned velocity deficit is shown in outer scaling highlighting the universality in the outer layer. 
% 
The logarithmic scaling of the streamwise component is encountered as straight blue lines in panels \textbf{(a)} and \textbf{(b)} where 
the extent of the logarithmic range increases with $\RE$ towards lower values of $z^-$ and higher values of $z^+$ depending on the scaling. 
Importantly, the logarithmic region is widening for increasing Reynolds number -- irespective of the scaling. 
% 
In this simple inner scaling, the spanwise velocity (which follows a mixed scaling) 
Does not collapse but seems to depend on $Re$ (the collapse is seen in Fig.~\ref{fig:inner_w}).
% 
However, the velocity deficit in outer units becomes approximately universal, also across the inner layer; 
this reflects the compensation of reduced turning ($\alpha$) by increased drag ($u_\star$), 
and is consistent with the theoretical considerations discussed in \ref{sec:drag-law}. 
% 
\par
%
The spanwise velocity at a fixed height scales approximately as $Re_\tau^{-1}$ (Sec.~\ref{sec:spanwise}).
%
However, the fraction  of turning that is encountered within the inner layer of the flow amounts to about 1/3 of the total wind veer  (Fig.~\ref{fig:etling}a).  
%
This is because, in inner units, the inner layer grows as $Re_\tau$  which exactly compensates the reduced gradient of spanwise velocity. 
% 
The hodographs show the well-known Ekman shape with the laminar profile as an outer limit and 'flatter'  hodographs, corresponding to less turning, for increasing Reynolds number.

@@

\subsection{Comparison with other theories}
%%% VAN DRIEST SCALING
\begin{figure}
  \centerline{\includegraphics[width=0.66\textwidth]{../plot/vanDriest_profile.png}}
  \caption{Near-wall velocity profile according to the 
    van-Driest scaling (blue, solid) in comparison with 
    the present model (red, dashed), 
    the viscous law of the wall (orange dashed), and 
    the logarithmic law (green, solid)} 
  \label{fig:vanDriest}
\end{figure}  An alternative approach that considers viscous effects close to the surface is the van-Driest scaling \citep{vandriest:JOT1956}, 
where an exponential damping of Prandtl's mixing length is considered near the wall to yield 
\begin{align}
  \frac{\p u^+}{\p z^+} = \frac{2}{ 1+ \sqrt{1+\left(2\kappa z^+\right)^2 (1-\exp\left[-z^+\right])}};
\end{align}
the spanwise component is zero as no rotational effects are considered. 
%
Comparing our proposed formulation for the stream-wise velocity in the inner layer to  \citeauthor{vandriest:JOT1956}'s formulation yields later convergence of the velocity onto the logarithmic profile while, over all, it serves as an excellent model of the streamwise velocity cmponent (Fig.~\ref{fig:vanDriest}):
% 
Notable deviations (on the order of few percent) only occur in the region $10 < z^+ < 30$,  where the velocity transitions from the linear to the logarithmic profile. 
%
\par
%
%EKMAN SPIRAL 
For the higher layers of the ABL, the Ekman spiral is the simplest model available. 
%
When employed across the entirety of the ABL, the spiral of a turbulent Ekman layer is flattened with respect to Ekman's laminar solution, which  corresponds to a reduction of the veering angle both at the surface and throughout the ABL. 
%
We, however, find that a modified version of the Ekman spiral explicitly taking into account the surface boundary condition, is a consistent model and yields excellent agreement with the velocity profiles from DNS (Sec.~\ref{sec:ekman}).

@@

A two-layer model consolidating both the logartihmic and Ekman layer can be obtained following the arguments by \cite{etling:2008}, cf. \cite{emeis:2018}. Given a surface veering and a matching height (extent of the logarithmic layer), a formulation for the velocity profile across both the logarithmic and the outer layer is obtained. A comparison using the surface veering based on our model and a matching height of $z_\text{prandtl}=0.05\delta$, which gives better results than the matching height of $0.1\delta$ suggested by \cite{etling:2008}, is shown in Fig.~\ref{fig:etling}. 

>>>>

The Coriolis force bends the apparent path of motion on a rotating sphere and establishes geostrophic equilibrium when in balance with a pressure gradient force. Wind veer away from the wind direction in geostrophic equilibrium is
(i) due to direct frictional effects in the very vicinity of the surface and
(ii) due to turbulence which exerts indirect frictional effects; these effects
cause a slow-down of the mean wind reducing the Coriolis force thus turning the wind
in favor of the pressure gradient force. 
%%
Not only does the veering set the frame of reference for surface layer theory, it also has
effects at small and large scales from large-scale dispersion via plume spreading to
cyclone spin-down \citep{svensson:BM2009}
and on the capabilities of data assimilation and accuracy of surface flux estimates \citep{brown:QJR2005}. 
%
From a large-scale perspective, the veering of wind across the planetary boundary layer determines
the amount of cross-isobaric mass-flux, commonly referred to as 'Ekman pumping' \citep{ekman:AMA1905},
and it is thus a key factor in the life-cycle of large-scale synoptic systems.
%
% %% This shear is probably not related to Ekman-turning 
% On the meso-scale, shear in the surface layer a crucial factor for the initiation and sustained strong convective events (REFERENCE?).
% 
Within the atmospheric boundary layer (ABL), directional shear of the wind in
the upper part of the surface layer may cause a systematic yaw for tall wind power generation devices
where blades reach into the Ekman layer, i.e. that part of the boundary layer where the wind starts to turn;
an exact estimate of such effects is critical in the site assessments for wind farms
\citep{calaf:PF2010, mirocha:WES2018}. 
%
\par
%
In the planetary boundary layer, wind veer is characterized by the surface veering angle $\alpha$
defined as the angle between the negative surface shear stress $\tau_\SFC$ and the geostrophic wind.
%
Surface veering $\alpha$ and geostrophic drag $Z\equiv u_\star/G$, where the friction velocity $u_\star\equiv\sqrt{|\tau_\SFC|/\rho}$,
uniquely determine the surface drag $\tau_\SFC$ in a turbulent Ekman flow. 
%
In any quantitative description of the surface layer, the friction velocity $u_\star$ is the dynamic scale
and $\alpha$ defines the horizontal alignment of the frame of reference, i.e. the rotation of the surface 
friction with respect to the outer wind direction. 
%
Knowledge about $u_\star$ and $\alpha$ is thus a prerequisite for any quantitative theory of the surface layer, 
and \citet{rossby:PIP1935} constrained the two parameters based on integral relations in the ABL.
%
Asymptotic similarity theory was later used by \cite{tennekes:JAS1973, blackadar:JAS1968},
and--based on his seminal direct numerical simulations (DNS) of Ekman flow--,~\cite{spalart:JFM1989} suggested
a modification to take into account effects of low to intermediate Reynolds numbers.
Later on, constants were re-evaluated with a focus on the ABL based on
observations \citep{hogstrom:BM1988,hogstrom:BM1996}
and numerical modelling \citep{spalart:PF2008, spalart:PF2009, ansorge:BM2014, ansorge:BM2019}.
%
\par
%
Attempts were also undertaken to obtain profiles of the wind speed: 
One approach is to match the inner and outer layer at a reference height as suggested by
\cite{etling:2002, emeis:2018}  (Sec. 21.10; Eq. 21.48); they choose the Prandtl-layer height $z_\text{Prandtl}$ to match
the wind speed profiles, which, however, requires external prescription of $\alpha(z_\text{prandtl})$, the veering at that height. 
A one-dimensional profile with constant veering is given by \citet[Sec. 3; Eq. 3.1-3.19]{emeis:m2007}. 
%%%

@@

We therefore resort to the physical manifestation of the eddy diffusivity in an Ekman layer, and use the

@@

This can, in particular for low or extremely high $\RE$, cause discontinuity in the derivatives of the velocity profiles around the transition from the inner to the outer layer. 

@@

height $z_{blend}^-=0.28-2.25\sqrt{1/Re_\tau}$ and a transition thickness 

@@

  \omega(z^-) = \frac12 \left[\text{erf}\left(z_\text{trans}\log\frac{z^-}{z_{\text{blend}}^{-}}\right) +1\right],

@@

%
\begin{figure}
  \phantom{AAA}\textbf{(a)\hspace{0.3\textwidth}(b)\hspace{0.3\textwidth}(c)}\\
  \includegraphics[width=\textwidth]{../plot/ekman_profiles.pdf}
  \caption{ \textbf{(a)} Velocity deficit, \textbf{(b)} velocity profile in shear-aligned hodographs and \textbf{(c)} hodograph in geostrophy-aligned coordinates.
   In panels \textbf{(a)} and \textbf{(b)}, blue lines correspond to the streamwise component and red to the spanwise. }
  \label{fig:summary} 
\end{figure} 
The resulting velocity profiles across the entire boundary layer are shown in inner and outer scaling as well as in ho\-do\-graph- view in Fig.~\ref{fig:summary}. The shear-aligned velocity deficit is shown in outer scaling highlighting the universality in the outer layer. 
% 
The logarithmic scaling of the streamwise component is encountered as straight blue lines in panels \textbf{(a)} and \textbf{(b)} where 
the extent of the logarithmic range increases with $\RE$ towards lower values of $z^-$ and higher values of $z^+$ depending on the scaling. 
Importantly, the logarithmic region is widening for increasing Reynolds number -- irrespective of the scaling. 
% 
In this simple inner scaling, the spanwise velocity (which follows a mixed scaling) 
does not collapse but seems to depend on $Re$ (the collapse is seen in Fig.~\ref{fig:inner_w}).
% 
However, the velocity deficit in outer units becomes approximately universal, also across the inner layer; 
this reflects the compensation of reduced turning ($\alpha$) by increased drag ($u_\star$), 
and is consistent with the theoretical considerations discussed in \ref{sec:drag-law}. 
% 
\par
%
The spanwise velocity at a fixed height scales approximately as $Re_\tau^{-1}$ (Sec.~\ref{sec:spanwise}).
%
However, the fraction  of turning that is encountered within the inner layer of the flow amounts to about 1/3 of the total wind veer  (Fig.~\ref{fig:etling}a).  
%
This is because, in inner units, the inner layer grows as $Re_\tau$  which exactly compensates the reduced gradient of spanwise velocity. 
% 
The hodographs show the well-known Ekman shape with the laminar profile as an outer limit and 'flatter'  hodographs, corresponding to less turning, for increasing Reynolds number.

@@

\subsection{Comparison with other theories}
%%% VAN DRIEST SCALING
\begin{figure}
  \centerline{\includegraphics[width=0.66\textwidth]{../plot/vanDriest_profile.png}}
  \caption{Near-wall velocity profile according to the 
    van-Driest scaling (blue, solid) in comparison with 
    the present model (red, dashed), 
    the viscous law of the wall (orange dashed), and 
    the logarithmic law (green, solid)} 
  \label{fig:vanDriest}
\end{figure}  An alternative approach that considers viscous effects close to the surface is the van-Driest scaling \citep{vandriest:JOT1956}, 
where an exponential damping of Prandtl's mixing length is considered near the wall to yield 
\begin{align}
  \frac{\p u^+}{\p z^+} = \frac{2}{ 1+ \sqrt{1+\left(2\kappa z^+\right)^2 (1-\exp\left[-z^+\right])}};
\end{align}
the spanwise component is zero as no rotational effects are considered. 
%
Comparing our proposed formulation for the stream-wise velocity in the inner layer to  \citeauthor{vandriest:JOT1956}'s formulation yields later convergence of the velocity onto the logarithmic profile while, over all, it serves as an excellent model of the streamwise velocity component (Fig.~\ref{fig:vanDriest}):
% 
Notable deviations (on the order of few percent) only occur in the region $10 < z^+ < 30$,  where the velocity transitions from the linear to the logarithmic profile. 
%
\par
%
%EKMAN SPIRAL 
For the higher layers of the ABL, the Ekman spiral is the simplest model available. 
%
When employed across the entirety of the ABL, the spiral of a turbulent Ekman layer is flattened with respect to Ekman's laminar solution, which  corresponds to a reduction of the veering angle both at the surface and throughout the ABL. 
%
We, however, find that a modified version of the Ekman spiral explicitly taking into account the surface boundary condition, is a consistent model and yields excellent agreement with the velocity profiles from DNS (Sec.~\ref{sec:ekman}).

@@

A two-layer model consolidating both the logarithmic and Ekman layer can be obtained following the arguments by \cite{etling:2008}, cf. \cite{emeis:2018}. Given a surface veering and a matching height (extent of the logarithmic layer), a formulation for the velocity profile across both the logarithmic and the outer layer is obtained. A comparison using the surface veering based on our model and a matching height of $z_\text{prandtl}=0.05\delta$, which gives better results than the matching height of $0.1\delta$ suggested by \cite{etling:2008}, is shown in Fig.~\ref{fig:etling}. 
